---
title: "Analysis of TCGA Lung Squamous Cell Carcinoma (LUSC)"
author: "Estefania Mancini"
date: "May 2020"
output:
  html_document:
    code_folding: hide
    fig_caption: yes
    fig_height: 6
    fig_width: 6
    number_sections: yes
    toc: yes
  pdf_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, message=FALSE, warning=FALSE, results='hide'}
library(ggplot2)
library(survival)
library(knitr)
library(kableExtra)
library(survminer)
library(ggfortify)
library(psichomics)
library(xtable)
```

# Introduction
The Cancer Genome Atlas (TCGA, https://www.cancer.gov/about-nci/organization/ccg/research/structural-genomics/tcga), molecularly characterized over 20,000 primary cancer and matched normal samples spanning 33 cancer types. 

# Data
* Raw data is available under folder <"data">

* Serrano’s group shared with us the tables of the splicing events for LUNG cancer produced using (https://github.com/vastgroup/vast-tools). We extratcted PSI for NUMB EXON 9 (HsaEX0044216) as well as coversion tables for samples and clinical data (see <"../data/psi_lung_NA_NewN3.tab"/)

* Clinical data was obtained using psichomics (see https://www.bioconductor.org/packages/release/bioc/html/psichomics.html). See <../data/luad_clinical_psichomics.tab>.

* nrow=503 (samples)
* ncol=2169 (features)
* adin.disease_code=LUSC


```{r clinical, echo=FALSE, message=FALSE, warning=FALSE, results='hide'}
clinical<-read.table( "../data/lusc_clinical_psichomics.tab", sep="\t", header = T, row.names = 1)
cancertype="LUSC"
#kable(clinical[1:3, 1:5], caption = "CLinical data")
gender<-grep("patient.gender", 
             colnames(clinical))
vitalStatus<-grep("patient.vital_status", 
                  colnames(clinical))
sample_type<-grep("patient.samples.sample.2.sample_type", colnames(clinical))
age<-grep("patient.age_at_initial_pathologic_diagnosis",colnames(clinical))
# mutations 
muts<-grep("mutation", colnames(clinical) )
colnames(clinical)[muts]
EGFR<-grep("EGFR",colnames(clinical),ignore.case = TRUE)
KRAS<-grep("KRAS",colnames(clinical),ignore.case = TRUE)
alk<-grep("alk", colnames(clinical),ignore.case = TRUE)
#Check info for translocation as well:
trans<-grep("translocation", colnames(clinical)); 
colnames(clinical)[trans]
# Ientify columns for tumor description:
stage<-grep("patient.stage_event.pathologic_stage_tumor_stage", colnames(clinical))
colnames(clinical)[stage]
neoplasm_cancer_status<-grep("patient.person_neoplasm_cancer_status", colnames(clinical))
colnames(clinical)[neoplasm_cancer_status]

# Define the columns for analysis of survival:
follow_up.vital_status<-grep("patient.follow_ups.follow_up.vital_status", colnames(clinical))
days_to_death<-grep("patient.days_to_death", colnames(clinical))
days_last_follow<-grep("patient.days_to_last_followup", colnames(clinical))
days_last_known_alive<-grep("patient.days_to_last_known_alive", colnames(clinical))
fup2<-grep(
  "patient.follow_ups.follow_up.2.days_to_new_tumor_event_after_initial_treatment",
  colnames(clinical))
fup3<-  grep(
  "patient.follow_ups.follow_up.3.days_to_new_tumor_event_after_initial_treatment",
           colnames(clinical))
fup4<-grep(
  "patient.follow_ups.follow_up.4.days_to_new_tumor_event_after_initial_treatment",
    colnames(clinical))
newTumourI<-grep(
  "patient.follow_ups.follow_up.days_to_new_tumor_event_after_initial_treatment", 
  colnames(clinical))
newTumourE<-grep(
  "patient.new_tumor_events.new_tumor_event.days_to_new_tumor_event_after_initial_treatment",
  colnames(clinical))

canonical_status<-grep("patient.bcr_canonical_check.bcr_patient_canonical_status", colnames(clinical))
treatment_success<-grep("patient.follow_ups.follow_up.followup_treatment_success", colnames(clinical))
therapy_outcome_success<-grep("patient.follow_ups.follow_up.primary_therapy_outcome_success", colnames(clinical))
primary_therapy_outcome_success<-grep("patient.primary_therapy_outcome_success", colnames(clinical))
smoker<-grep("patient.number_pack_years_smoked", colnames(clinical)) 
#We made a small datframe for Recurrence Free Survival (RFS):
RFS<-data.frame(clinical$patient.days_to_death, 
                clinical$patient.days_to_last_followup, 
                clinical$patient.days_to_last_known_alive,
                clinical$patient.follow_ups.follow_up.2.days_to_new_tumor_event_after_initial_treatment,
                clinical$patient.follow_ups.follow_up.3.days_to_new_tumor_event_after_initial_treatment,
                clinical$patient.follow_ups.follow_up.4.days_to_new_tumor_event_after_initial_treatment,
                clinical$patient.follow_ups.follow_up.days_to_new_tumor_event_after_initial_treatment,
                clinical$patient.new_tumor_events.new_tumor_event.days_to_new_tumor_event_after_initial_treatment)
rownames(RFS)<-rownames(clinical)

#Make a small dataframe with our selected columns:
subset<-clinical[,unique(c(gender,
                  vitalStatus,
                  sample_type,
                  age,
                  muts,
                  EGFR,#2
                  KRAS,
                  alk,
                  trans,
                  stage,
                  neoplasm_cancer_status,
                  follow_up.vital_status,
                  days_to_death,
                  days_last_follow,
                  days_last_known_alive,
                  fup2,
                  fup3,
                  fup4,
                  newTumourI,
                  newTumourE,
                  canonical_status,
                  treatment_success,
                  therapy_outcome_success,
                  primary_therapy_outcome_success,
                  smoker))]
# Clean a bit column names:
colnames(subset)<-sub("patient.", "", colnames(subset))
subset$stage_event.pathologic_stage_tumor_stage<-sub("stage ","",
subset$stage_event.pathologic_stage_tumor_stage);
#Prepare a new column: *Aggregate stage*, where we classify stages as *early* (i, ia, ib, ii, iia, iib) o *late* (iii,iiia,iiib,iv)
subset$aggStage <- subset$stage
subset$aggStage[subset$aggStage=="i"]<-"early"
subset$aggStage[subset$aggStage=="ia"]<-"early"
subset$aggStage[subset$aggStage=="ib"]<-"early"
subset$aggStage[subset$aggStage=="ii"]<-"early"
subset$aggStage[subset$aggStage=="iia"]<-"early"
subset$aggStage[subset$aggStage=="iib"]<-"early"
subset$aggStage[subset$aggStage=="iii"]<-"late"
subset$aggStage[subset$aggStage=="iiia"]<-"late"
subset$aggStage[subset$aggStage=="iiib"]<-"late"
subset$aggStage[subset$aggStage=="iv"]<-"late"
```

```{r newVar, echo = TRUE,message=TRUE, warning=TRUE}
subset$newVar<-paste(subset$stage,
                     subset$gender, sep="_")

dfAg1<-data.frame(table(subset$newVar))
#Var1 viene del table
dfAg1<-cbind(dfAg1,   
             data.frame(matrix(unlist(strsplit(as.character(dfAg1$Var1), "_")), ncol=2, byrow = T)))

colnames(dfAg1)[1]<-"comb"
colnames(dfAg1)[2]<-"NumOfSamples"
colnames(dfAg1)[3]<-"stage"
colnames(dfAg1)[4]<-"gender"

knitr::kable(dfAg1[,2:4], format="pandoc", align='cccc')

ggplot(data=dfAg1, aes(x=stage, y=NumOfSamples, fill=gender)) +
  geom_bar(stat="identity",  position="dodge")+
  theme_minimal()

subset$newVar2<-paste(subset$aggStage,
                      subset$gender, 
                      sep="_")
dfAg2<-data.frame(table(subset$newVar2))
#Var1 viene del table
dfAg2<-cbind(dfAg2,   
             data.frame(matrix(unlist(strsplit(as.character(dfAg2$Var1), "_")), ncol=2, byrow = T)))

colnames(dfAg2)[1]<-"comb"
colnames(dfAg2)[2]<-"NumOfSamples"
colnames(dfAg2)[3]<-"aggStage"
colnames(dfAg2)[4]<-"gender"

knitr::kable(dfAg2, format="pandoc", align='cccc')

ggplot(data=dfAg2, aes(x=aggStage, y=NumOfSamples, fill=gender)) +
  geom_bar(stat="identity",  position="dodge")+
  theme_minimal()
```


# Descriptive 

Here we analyze PSI distribution using several clinical variables

## Statistical tests
We use:
  
* Wilcoxon test	wilcox.test()	Compare two groups (non-parametric)

* Kruskal-Wallis	kruskal.test()	Compare multiple groups (non-parametric)

See <http://www.sthda.com/english/articles/24-ggpubr-publication-ready-plots/76-add-p-values-and-significance-levels-to-ggplots/>
  
```{r psiVal, echo = FALSE, message=FALSE, warning=FALSE}
psiVal<-read.table("../data/psi_lung_NA_NewN3.tab", 
                   header = T, 
                   stringsAsFactors = F, 
                   row.names = 1)
#kable(psiVal[, 1:10], caption = "PSI data")
#
sIds<-colnames(psiVal)[7:ncol(psiVal)]
sIdsB<-gsub('\\.', '-',as.character(sIds))
sIdsB<-gsub('^X', '',as.character(sIdsB))
colnames(psiVal)<-gsub('\\.', '-',as.character(colnames(psiVal)))
colnames(psiVal)<-gsub('^X', '',as.character(colnames(psiVal)))
```


```{r convTable, echo = FALSE, message=FALSE, warning=FALSE}
## Load conversion tables, for sample ID, case:
pt<-read.table("../data/lung_pt.files.txt", header = T, sep="\t")
#kable(pt[1:5,], caption = "Conversion data")
convTablePT<-data.frame(case=pt$cases.0.submitter_id, file=pt$file_id)
convTablePT$origin<-rep("pt", nrow(convTablePT))
stn<-read.table("../data/lung_stn.files.txt", header = T, sep="\t")
convTableSTN<-data.frame(case=stn$cases.0.submitter_id, file=stn$file_id)
convTableSTN$origin<-rep("stn", nrow(convTableSTN))
ConvTotal<-rbind(convTablePT, convTableSTN)
psiNormali<-match(convTableSTN$file, colnames(psiVal))
psiN<-as.numeric(t(psiVal[, psiNormali]))
notna<-which(is.na(psiN))
psiNormal<-as.numeric(psiN[-notna])
g0Max<-max(psiNormal)
g0Mean<-mean(psiNormal)
dfNormal<-data.frame(psiNormal, rep("normal",length(psiNormal)))
colnames(dfNormal)<-c("psi", "type")
#Match samples PT:
ii<-match(sIdsB, ConvTotal$file)
pp<-match(rownames(subset), ConvTotal$case)
ppi<-pp[!is.na(pp)]
ppo<-which(!is.na(pp))
finalDF<-cbind(subset[ppo,], ConvTotal[ppi,])
finalDF<-finalDF[finalDF$origin!="stn",]
#Agregamos el PSI
vv<-match(finalDF$file, colnames(psiVal))
values<-psiVal[,vv]
values[values=="NAold"]<-NA
values[values=="NAnew3"]<-NA
finalV<-as.numeric(t(values ))
names(finalV)<-colnames(values)
#Armo este DF solo para plotear Tumor vs Normal:
dfCancer<-data.frame(finalV, rep("tumor", length(finalV) )) 
colnames(dfCancer)<-c("psi", "type")
dfNormal<-data.frame(psiNormal, 
                     rep("normal",length(psiNormal)))
colnames(dfNormal)<-c("psi", "type")
dfPlots<-rbind(dfCancer, dfNormal)
isna<-which(is.na(dfPlots$psi))
dfPlots<-dfPlots[-isna,]
dfPlots$color<-as.character(dfPlots$type)
dfPlots$color[dfPlots$color=="tumor"]<-"tomato"
dfPlots$color[dfPlots$color=="normal"]<-"skyblue"
dfPlots$color<-as.factor(dfPlots$color)
```

## Analysis by type: normal or tumoral tissue (samples)

For the first analysis we compared the levels of NUMB exon 9 inclusion between healthy and tumoral tissue.
In total 109 healthy samples (mean PSI 6.14) vs  494 tumoral samples (mean PSI 35) were compared using a Wilcoxon test.

This data suggests that lung tumors have a (significant) higher NUMB exon 9 PSI value than healthy tissue (p-value < 2.2e-16).


```{r violinB, echo = TRUE,message=TRUE, warning=TRUE}
normal<-dfPlots[dfPlots$type=="normal",]
print(paste("N normal:",dim(normal)[1]))
print(paste("Mean normal:",mean(normal$psi)))
tumor<-dfPlots[dfPlots$type=="tumor",]
print(paste("N tumoral:",dim(tumor) [1]))
print(paste("Mean tumoral:",mean(tumor$psi)))
wilcox.test(normal$psi, tumor$psi)
```

```{r violinC, echo = TRUE,message=FALSE, warning=FALSE}
numb_psi<-dfPlots[,1:2]
colnames(numb_psi) <- c("PSI","type")
ggplot(data= numb_psi, aes(x=type,y=PSI)) + geom_violin(aes(fill=type)) +
  scale_colour_manual(values = c("skyblue", "tomato")) +  
  scale_fill_manual(values = c("skyblue", "tomato")) + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1))+
  geom_jitter(shape=16, 
              position=position_jitter(0.05),
              size=.5, 
              color=  "black")+
  geom_boxplot(width=0.1) +
  stat_summary(fun=median, geom="point", size=1, color="red")+   stat_compare_means(label.y = 100, label.x=2)+
  theme_minimal()
```

### GTEX (caterina)
We also analyzed PSI value in healthy tissues, using The Genotype-Tissue Expression (GTEx) project data (https://www.gtexportal.org/home/)

* Vast-tool hg19,PSI range: 0-100
* median PSI values across lung samples = 3.52%
* mean PSI values across lung samples= 3.52%
* Range PSI values across lung samples= 20%

This data suggest PSI in healthy tissues is ~3.5 %

## Analysis by stage

We analyzed PSI values according to their tumoral stage (provided in the clinical data) using a kruskal.test(). We didn't find any significant differences between them according kruskal test (pval = 0.24)

```{r stage, echo = FALSE, message=FALSE, warning=FALSE}
df<-data.frame(sample=names(finalV),psi=as.numeric(finalV))
cc<-match(df$sample, finalDF$file)
finalV<-data.frame(df, finalDF[cc,])
finalV<-finalV[-which(is.na(finalV$psi)),]
finalV$status<-rep(1,  nrow(finalV))
isna<-which(is.na(finalV$stage))
finalVNotNA<-finalV[-isna,]
print(table(finalVNotNA$stage))
finalVNotNA$stage<-as.factor(finalVNotNA$stage)
kruskal.test(psi ~ stage, data = finalVNotNA)
```

```{r stageP, echo = FALSE, message=FALSE, warning=FALSE}
ggplot(finalVNotNA, aes(x=stage, y=psi, fill=stage))+ 
  geom_violin()   + 
  geom_jitter(shape=16, 
              position=position_jitter(0.05),
              size=.5, 
              color=  "black")+
  geom_boxplot(width=0.1) +
  stat_summary(fun=median, geom="point", size=1, color="red")+   stat_compare_means(label.y = 100, label.x=2)+
  theme_minimal()
```

* (Now, re plot ordering by median, to improve visualization):

```{r stagePorder, echo = FALSE, message=FALSE, warning=FALSE}
finalVNotNA$stage = with(finalVNotNA, reorder(stage, psi, median))

ggplot(finalVNotNA, aes(x=stage, y=psi, fill=stage))+ 
  geom_violin()   + 
  geom_jitter(shape=16, 
              position=position_jitter(0.05),
              size=.5, 
              color=  "black")+
  geom_boxplot(width=0.1) +
  stat_summary(fun=median, geom="point", size=1, color="red")+   stat_compare_means(label.y = 100, label.x=2)+
  theme_minimal()
```

* **Results**: According Kruskal-Wallis ( p-value = 0.25) there is no clear stage with (high) different psi. 

## Analysis by Aggregate stage
In order to have bigger groups, we aggregated the tumors in two main stages.
 Early (n=400, stage I to IIIB) and Late (n=90, stage III and IV). We compared NUMB exon 9 PSI in these groups.

* **Results**: despite Late stage seems to has higher PSI, there is not a significant difference according to Wilcoxon test p-value = 0.1856)

```{r aggStageWilcoxon, echo = FALSE, message=FALSE, warning=FALSE}
print(table(finalVNotNA$aggStage))
wilcox.test(finalVNotNA$psi[finalVNotNA$aggStage=="early"],
            finalVNotNA$psi[finalVNotNA$aggStage=="late"])
```


```{r aggStage, echo = FALSE, message=FALSE, warning=FALSE}
ggplot(finalVNotNA, aes(x=aggStage, y=psi, fill=aggStage))+ 
  geom_violin()   + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1))+
  geom_jitter(shape=16, 
              position=position_jitter(0.05),
              size=.5, 
              color=  "black")+
  geom_boxplot(width=0.1) +
  stat_summary(fun=median, geom="point", size=1, color="red")+   stat_compare_means(label.y = 100, label.x=2)+
  theme_minimal()
```


## Analysis by gender
 
We compared NUMB exon 9 PSI in Male (n=363 vs Female n=128) and also in Early and Late tumors. 

* **Results**
We didnt find signtificant difference in PSI between male and female samples ( p-value = 0.8514)

```{r genderwilcoxon, echo = FALSE, message=FALSE, warning=FALSE}
table(finalVNotNA$gender)
wilcox.test(finalVNotNA$psi[finalVNotNA$gender=="male"],
            finalVNotNA$psi[finalVNotNA$gender=="female"])
```


```{r gender, echo = FALSE, message=FALSE, warning=FALSE}
ggplot(finalVNotNA, aes(x=gender, y=psi, fill=gender))+ 
  geom_boxplot(width=0.1) +
  geom_violin()   + 
  geom_jitter(shape=16, 
              position=position_jitter(0.05),
              size=.5, 
              color=  "black")+
  stat_summary(fun=median, geom="point", size=1, color="red")+           stat_compare_means(label.y = 100, label.x=2)+
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

## Analysis by gender and aggregate stage
* **Results**: there is differences at both stages according KW test (early, pval=0.01, late, pval=0.037).
But there is not significance using Wilcoxon test (early pval=0.91, late pval=0.84)

```{r genderAgg3tests, echo = FALSE, message=FALSE, warning=FALSE}
early<-finalVNotNA[finalVNotNA$aggStage=="early", ]
print("Early by gender:")
print(table(early$gender))
late<-finalVNotNA[finalVNotNA$aggStage=="late", ]
print("Late by gender:")
print(table(late$gender))
print("Kruskal-Wallis test:")
kruskal.test(psi ~ gender, data = early)
kruskal.test(psi ~ gender, data = late)
print("Wilcoxon test:")
wilcox.test(early$psi[early$gender=="male"],
            early$psi[early$gender=="female"])
wilcox.test(late$psi[late$gender=="male"],
            late$psi[late$gender=="female"])
```

```{r genderAgg3, echo = FALSE, message=FALSE, warning=FALSE}
ggplot(finalVNotNA, aes(x=aggStage, y=psi, fill=gender))+ 
  #geom_violin()   + 
  geom_boxplot(width=0.1) +
  geom_jitter(shape=16, 
              position=position_jitter(0.05),
              size=.5, 
              color=  "black")+
  stat_summary(fun=median, geom="point", size=1, color="red")+    stat_compare_means(label.y = 100, label.x=2)+
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

## Correlation between NUMB exon 9 PSI and known LUAD driver mutations

Some of the most known cancer-related mutations are also clinically measured in the patients. We performed a correlation analysis to try to identify if there is any correlation between NUMB exon 9 inclusion levels and mutations in known genes.

We found reported mutations in these genes:  

* We found reported mutations in these genes: 
  + KRAS, EGFR and ALK
* Mutations on these genes were not reported: 
  + HER2, BRAF,DDR2,AKT,FGFR,NOTCH,TP53,CDH10,NFE2L2,KMT2C,CDKN2A,FAT1 and PI3KCA


* ALK:	eml4 alk translocation performed: 485 no, 9 yes
+ results: 4 fish

* not reported: 
HER2, BRAF, ALK, DDR2, FGFR, NOTCH, TP53, CDH10, NFE2L2, KMT2C, CDKN2A, FAT1, PI3KCA

### *egfr*
Samples with mutation analysis performed: 
* YES: 21,
  + l858 1
  + ns 19
  + other  1

* NO:  473 (NAs were replaced by "NO") 

* (1 yes vs no) There is not statistically significant differences between these 2 groups (p-value = 0.2643). 
  
```{r egfr, echo = FALSE, message=FALSE, warning=FALSE}
finalV$egfr_mutation_performed[is.na(finalV$egfr_mutation_performed)] <-"no"
print(table(finalV$egfr_mutation_performed))
kruskal.test(finalV$psi ~ finalV$egfr_mutation_performed)

finalV$egfr_mutation_result<-as.character(finalV$egfr_mutation_result)
finalV$egfr_mutation_result[is.na(finalV$egfr_mutation_result)]<-"ns"

finalV$egfr_mutation_performed[is.na(finalV$egfr_mutation_performed)]<-"no"
finalV$egfr_mutation_result<-as.character(finalV$egfr_mutation_result)
finalV$egfr_mutation_result[is.na(finalV$egfr_mutation_result)]<-"ns"
finalV$egfr_mutation_performed = with(finalV, reorder(egfr_mutation_performed, psi, median))
finalVE<-finalV[finalV$egfr_mutation_performed=="yes",]

############yes or not
print(table(finalVE$egfr_mutation_result))
kruskal.test(finalVE$psi ~ finalVE$egfr_mutation_result)
```

* (2 yes) There is not statistically significant differences  (p-value = 0.6113). 
```{r egfrp0, echo = TRUE, message=FALSE, warning=FALSE}
ggplot(finalV, aes(x=egfr_mutation_performed, y=psi, fill=egfr_mutation_performed))+ 
  geom_boxplot()   + 
  #geom_jitter(shape=16,                position=position_jitter(0.05), size=.5,  color=  "black")+
  stat_summary(fun=median, geom="point", size=1, color="red")+     stat_compare_means(label.y = 100, label.x=2)+
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

* (2, only yes) Mutation results (reorder by median PSI):  
  
```{r egfrpeResult, echo = TRUE, message=FALSE, warning=FALSE}
finalVE$egfr_mutation_result = with(finalVE, reorder(egfr_mutation_result, psi, median))

ggplot(finalVE, aes(x=egfr_mutation_result, y=psi, fill=egfr_mutation_result))+ 
  geom_boxplot()   + 
  #geom_jitter(shape=16,                position=position_jitter(0.05), size=.5,  color=  "black")+
  stat_summary(fun=median, geom="point", size=1, color="red")+    stat_compare_means(label.y = 100, label.x=2)+
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

### *kras*
* KRAS mutation performed: 493 no, 1 yes -> other

### *alk*
* **Results**: We found 9 cases where Eml4_alk_translocation was performed (4 fish)

## Treatment Success

We also explored the PSI in different groups according their treatment success:
  
### (1) Follow up: follow up treatment success

* Variable name: follow_ups.follow_up.followup_treatment_success
* **Results**: Despite we didnt find significant differences (pval 0.19), it seems PSI is slighly higher at progressive stages.

```{r treatment1, echo = TRUE, message=FALSE, warning=FALSE}
finalV$follow_ups.follow_up.followup_treatment_success<-
  as.character(finalV$follow_ups.follow_up.followup_treatment_success)

finalV$follow_ups.follow_up.followup_treatment_success[is.na(finalV$follow_ups.follow_up.followup_treatment_success)]<-"unknown"


print(
  table(finalV$follow_ups.follow_up.followup_treatment_success))
##########################################################
#remuevo los unknown
clean<-finalV[finalV$follow_ups.follow_up.followup_treatment_success!="unknown",]
#remuevo los partial remission response 
clean<-clean[clean$follow_ups.follow_up.followup_treatment_success!="partial remission/response",]
dim(clean)#234
table(clean$follow_ups.follow_up.followup_treatment_success)

clean$follow_ups.follow_up.followup_treatment_success<-
  factor(clean$follow_ups.follow_up.followup_treatment_success,
         c("complete remission/response",
           "stable disease","progressive disease"))
###########################################################
```

```{r treatment2, echo = TRUE, message=FALSE, warning=FALSE}
ggplot(finalV, 
       aes(x=follow_ups.follow_up.followup_treatment_success, 
           y=psi,
           fill=follow_ups.follow_up.followup_treatment_success))+ 
  geom_violin()   + 
  theme_bw()+
  geom_jitter(shape=16, 
              position=position_jitter(0.2),
              size=.5, color="darkgrey")     + 
  stat_summary(fun=median, geom="point", size=1, color="red")  + 
  stat_compare_means(label.y = 100, label.x=2)+
  theme(legend.position="none") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

```{r treatment3, echo = TRUE, message=FALSE, warning=FALSE}

ggplot(clean, 
       aes(x=follow_ups.follow_up.followup_treatment_success, 
           y=psi,
           fill=follow_ups.follow_up.followup_treatment_success))+ 
  geom_violin()   + 
  theme_bw()+
  geom_jitter(shape=16, 
              position=position_jitter(0.2),
              size=.5, color="darkgrey")     + 
  stat_summary(fun=median, geom="point", size=1, color="red")  + 
  stat_compare_means(label.y = 100, label.x=2)+
  theme(legend.position="none") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

* (1b) Repito reorder by median PSI:
  
```{r treatment3reorder, echo = TRUE, message=FALSE, warning=FALSE}
clean$follow_ups.follow_up.followup_treatment_success = with(clean, reorder(follow_ups.follow_up.followup_treatment_success, psi, median))

ggplot(clean, 
       aes(x=follow_ups.follow_up.followup_treatment_success, 
           y=psi,
           fill=follow_ups.follow_up.followup_treatment_success))+ 
  geom_violin()   + 
  theme_bw()+
  geom_jitter(shape=16, 
              position=position_jitter(0.2),
              size=.5, color="darkgrey")     + 
  stat_summary(fun=median, geom="point", size=1, color="red")  + 
  stat_compare_means(label.y = 100, label.x=2)+
  theme(legend.position="none") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```


### (2) follow_ups.follow_up.primary_therapy_outcome_success :
  
* Variable: follow_ups.follow_up.primary_therapy_outcome_success 
* **Results**:  there is a significant difference according KW test (pval 0.017)

```{r treatment4, echo = TRUE, message=FALSE, warning=FALSE}
ggplot(finalV, 
       aes(x=follow_ups.follow_up.primary_therapy_outcome_success, 
           y=psi,
           fill=follow_ups.follow_up.primary_therapy_outcome_success))+ 
  geom_violin()   + 
  theme_bw()+
  geom_jitter(shape=16, 
              position=position_jitter(0.2),
              size=.5, color="darkgrey")     + 
  stat_summary(fun=median, geom="point", size=1, color="red")  + 
  stat_compare_means(label.y = 100, label.x=2)+
  theme(legend.position="none") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

```{r treatment5, echo = TRUE, message=FALSE, warning=FALSE}
table(clean$follow_ups.follow_up.primary_therapy_outcome_success)
#remuevo los unknown
clean<-finalV[!is.na(finalV$follow_ups.follow_up.primary_therapy_outcome_success),]
clean<-clean[clean$follow_ups.follow_up.primary_therapy_outcome_success!="partial remission/response",]
dim(clean)#234
table(clean$follow_ups.follow_up.primary_therapy_outcome_success)

ggplot(clean, 
       aes(x=follow_ups.follow_up.primary_therapy_outcome_success, 
           y=psi,
           fill=follow_ups.follow_up.primary_therapy_outcome_success))+ 
  geom_violin()   + 
  theme_bw()+
  geom_jitter(shape=16, 
              position=position_jitter(0.2),
              size=.5, color="darkgrey")     + 
  stat_summary(fun=median, geom="point", size=1, color="red")  + 
  stat_compare_means(label.y = 100, label.x=2)+
  theme(legend.position="none") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

* (2b) Repito clean and reorder:
  + **Results** There is a clear difference, high PSI at more progressive stages (KW test, pval 0.0062)


```{r treatment5reorder, echo = TRUE, message=FALSE, warning=FALSE}
clean$follow_ups.follow_up.primary_therapy_outcome_success = with(clean, reorder(follow_ups.follow_up.primary_therapy_outcome_success, psi, median))
table(clean$follow_ups.follow_up.primary_therapy_outcome_success)
ggplot(clean, 
       aes(x=follow_ups.follow_up.primary_therapy_outcome_success, 
           y=psi,
           fill=follow_ups.follow_up.primary_therapy_outcome_success))+ 
  geom_violin()   + 
  theme_bw()+
  geom_jitter(shape=16, 
              position=position_jitter(0.2),
              size=.5, color="darkgrey")     + 
  stat_summary(fun=median, geom="point", size=1, color="red")  + 
  stat_compare_means(label.y = 100, label.x=2)+
  theme(legend.position="none") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```


### Primary therapy outcome success
* variable: primary_therapy_outcome_success
* **Results**: It seems that PSI values correlates with the progression of the disease, but we didnt find significant differences
+  complete remission/response
+  stable disease 
+  progressive disease 


```{r treatment6, echo = TRUE, message=FALSE, warning=FALSE}
finalV$primary_therapy_outcome_success<-as.character(finalV$primary_therapy_outcome_success)

finalV$primary_therapy_outcome_success[is.na(finalV$primary_therapy_outcome_success)]<-"unknown"
table(finalV$primary_therapy_outcome_success)
ggplot(finalV, 
       aes(x=primary_therapy_outcome_success, 
           y=psi,
           fill=primary_therapy_outcome_success))+ 
  geom_violin()   + 
  theme_bw()  +
  geom_jitter(shape=16, 
              position=position_jitter(0.2),
              size=.5, color="darkgrey")     + 
  stat_summary(fun=median, geom="point", size=1, color="red")  + 
  stat_compare_means(label.y = 100, label.x=2)+
  theme(legend.position="none") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```


```{r treatment6Clean, echo = TRUE, message=FALSE, warning=FALSE}
finalV$primary_therapy_outcome_success<-as.character(finalV$primary_therapy_outcome_success)

finalV$primary_therapy_outcome_success[is.na(finalV$primary_therapy_outcome_success)]<-"unknown"
table(finalV$primary_therapy_outcome_success)
clean<-finalV[finalV$primary_therapy_outcome_success!="unknown",]
clean<-clean[clean$primary_therapy_outcome_success!="partial remission/response",]
clean<-clean[!is.na(clean$primary_therapy_outcome_success),]
table(clean$primary_therapy_outcome_success)
ggplot(clean, 
       aes(x=primary_therapy_outcome_success, 
           y=psi,
           fill=primary_therapy_outcome_success))+ 
  geom_violin()   + 
  theme_bw()  +
  geom_jitter(shape=16, 
              position=position_jitter(0.2),
              size=.5, color="darkgrey")     + 
  stat_summary(fun=median, geom="point", size=1, color="red")  + 
  stat_compare_means(label.y = 100, label.x=2)+
  theme(legend.position="none") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

* (3b) repito order

```{r treatment6CleanOrder, echo = TRUE, message=FALSE, warning=FALSE}
table(clean$primary_therapy_outcome_success)

clean$primary_therapy_outcome_success = with(clean, reorder(primary_therapy_outcome_success, psi, median))

ggplot(clean, 
       aes(x=primary_therapy_outcome_success, 
           y=psi,
           fill=primary_therapy_outcome_success))+ 
  geom_violin()   + 
  theme_bw()  +
  geom_jitter(shape=16, 
              position=position_jitter(0.2),
              size=.5, color="darkgrey")     + 
  stat_summary(fun=median, geom="point", size=1, color="red")  + 
  stat_compare_means(label.y = 100, label.x=2)+
  theme(legend.position="none") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

## Neoplasm cancer status
We want to check if there were any difference according tumor types. There is not significant difference (WT, pval 0.9)

```{r treatment7, echo = TRUE, message=FALSE, warning=FALSE}
finalV$person_neoplasm_cancer_status<-as.character(finalV$person_neoplasm_cancer_status)
finalV$person_neoplasm_cancer_status[is.na(finalV$person_neoplasm_cancer_status)]<-"unknown"
print("Before cleaning")
table(finalV$person_neoplasm_cancer_status)
#remvove unknown
finalVNotUnknown<-finalV[finalV$person_neoplasm_cancer_status!="unknown",]
finalVNotUnknown<-finalVNotUnknown[!is.na(finalV$person_neoplasm_cancer_status),]
finalVNotUnknown<-finalVNotUnknown[finalVNotUnknown$person_neoplasm_cancer_status!="NA",]
print("After cleaning")
table(finalVNotUnknown$person_neoplasm_cancer_status)
ggplot(finalVNotUnknown, 
       aes(x=person_neoplasm_cancer_status, 
           y=psi,
           fill=person_neoplasm_cancer_status ))+ 
  geom_violin()   + 
  theme_bw()+
  geom_jitter(shape=16, 
              position=position_jitter(0.2),
              size=.5, color="darkgrey")     + 
  stat_summary(fun=median, geom="point", size=1, color="red")  + 
  stat_compare_means(label.y = 100, label.x=2)+
  theme(legend.position="none") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

## Sample types

```{r treatment8, echo = TRUE, message=FALSE, warning=FALSE}
finalV$samples.sample.2.sample_type<-as.character(finalV$samples.sample.2.sample_type)
finalV$samples.sample.2.sample_type[is.na(finalV$samples.sample.2.sample_type)]<-"unknown"
table(finalV$samples.sample.2.sample_type)

#order:
finalV$samples.sample.2.sample_type = with(finalV, reorder(samples.sample.2.sample_type, psi, median))

ggplot(finalV, 
       aes(x=samples.sample.2.sample_type, y=psi, fill=samples.sample.2.sample_type))+ 
  geom_violin()   + 
  theme_bw()+
  geom_jitter(shape=16, 
              position=position_jitter(0.2),
              size=.5, color="darkgrey")     + 
  stat_summary(fun=median, geom="point", size=1, color="red")  + 
  stat_compare_means(label.y = 100, label.x=2)+
  theme(legend.position="none") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

## Smokers
Explore if there is any relationship between PSI and Smoke. There are  417 patients which reported: "number pack years smoked"

```{r smoker, echo = TRUE, message=FALSE, warning=FALSE}
smoking<-finalV[!is.na(finalV$number_pack_years_smoked),]
dim(smoking)
plot(smoking$psi, smoking$number_pack_years_smoked)
cor(smoking$psi, smoking$number_pack_years_smoked)
lm(smoking$psi ~ smoking$number_pack_years_smoked)
```


# Survival Plots

To study the impact of alternative splicing events on prognosis, Kaplan-Meier curves may be plotted for groups of patients separated by a given PSI cutoff for a given alternative splicing event. The optimal PSI cutoff maximises the significance of group differences in survival analysis (i.e. minimises the p-value of the log-rank tests of difference in survival between individuals whose samples have a PSI below and above that threshold). 

## (1st strategy)  **optimalSurvivalCutoff**
We estimate optimo PSI using **psichomics optimalSurvivalCutoff** function. According to this function, **PSI =  50.83** is the value that best divide patients. We then defined:
  
* g0: PSI < 50.83: 396 
* g1: PSI > 50.83: 87  

Total number of cases: 483

```{r clinical0, echo = FALSE, message=FALSE, warning=FALSE}
#aca computamos el optimo sin remover NAs
clinical0<-data.frame(finalV$days_to_last_follow, 
                      finalV$days_to_death,
                      finalV$stage,
                      finalV$gender)
names(clinical0) <- c("patient.days_to_last_followup", 
                      "patient.days_to_death",
                      "patient.stage_event.pathologic_stage",
                      "patient.gender")
timeStart  <- "days_to_death"
event      <- "days_to_death"
psi<-as.numeric(finalV$psi)
opt <- optimalSurvivalCutoff(clinical0, 
                             psi, censoring="right", 
                             event="days_to_death", 
                             timeStart="days_to_death")
(optimalCutoff <- opt$par)
# Optimal exon inclusion level 69.40
(optimalPvalue <- opt$value)  # Respective p-value 0.00751
label     <- labelBasedOnCutoff(psi, round(optimalCutoff, 2), 
                                label="PSI values")
survTerms <- processSurvTerms(clinical0, censoring="right",
                              event="days_to_death", 
                              timeStart="days_to_death",
                              group=label, scale="years")
surv <- survfit(survTerms)
pvalue <- testSurvival(survTerms)
plotSurvivalCurves(surv, pvalue=pvalue, mark=FALSE)
```




```{r makeNotNAs, message=FALSE, warning=FALSE,  echo=FALSE}
survival<-finalV
#OS ##########################################################
NAs<-
  data.frame(survival$days_to_death,
             survival$days_to_last_followup,
             survival$days_to_last_known_alive);
SUMNAS<-rowSums(apply(NAs, 2,is.na) ); length(SUMNAS); survivalNotNas<-survival[SUMNAS!=3,]
survivalNotNas$days_to_death[is.na(survivalNotNas$days_to_death)]<-0
survivalNotNas$days_to_last_followup[is.na(survivalNotNas$days_to_last_followup)]<-0
survivalNotNas$days_to_last_known_alive[is.na(survivalNotNas$days_to_last_known_alive)]<-0
#
survivalNotNas$daysOS<-
  unlist(apply(data.frame(survivalNotNas$days_to_death,
                          survivalNotNas$days_to_last_followup,                      survivalNotNas$days_to_last_known_alive),
               1,max));
#RFS #########################################################
daysRFSDF<-
  data.frame(
    survivalNotNas$follow_ups.follow_up.2.days_to_new_tumor_event_after_initial_treatment,
    survivalNotNas$follow_ups.follow_up.3.days_to_new_tumor_event_after_initial_treatment,
    survivalNotNas$follow_ups.follow_up.4.days_to_new_tumor_event_after_initial_treatment,
    survivalNotNas$follow_ups.follow_up.days_to_new_tumor_event_after_initial_treatment,
    survivalNotNas$new_tumor_events.new_tumor_event.days_to_new_tumor_event_after_initial_treatment)
###################################################
SUMNASRFS<-rowSums(apply(daysRFSDF, 2,is.na));
survivalNotNas$follow_ups.follow_up.2.days_to_new_tumor_event_after_initial_treatment[
  is.na(survivalNotNas$follow_ups.follow_up.2.days_to_new_tumor_event_after_initial_treatment)]<-0
survivalNotNas$follow_ups.follow_up.3.days_to_new_tumor_event_after_initial_treatment[
  is.na(survivalNotNas$follow_ups.follow_up.3.days_to_new_tumor_event_after_initial_treatment)]<-0
survivalNotNas$follow_ups.follow_up.4.days_to_new_tumor_event_after_initial_treatment[
  is.na(survivalNotNas$follow_ups.follow_up.4.days_to_new_tumor_event_after_initial_treatment)]<-0
survivalNotNas$follow_ups.follow_up.days_to_new_tumor_event_after_initial_treatment[
  is.na(survivalNotNas$follow_ups.follow_up.days_to_new_tumor_event_after_initial_treatment)]<-0
survivalNotNas$new_tumor_events.new_tumor_event.days_to_new_tumor_event_after_initial_treatment[
  is.na(survivalNotNas$new_tumor_events.new_tumor_event.days_to_new_tumor_event_after_initial_treatment)]<-0
survivalNotNas$daysRFS<-
  unlist(apply(data.frame(
    survivalNotNas$follow_ups.follow_up.2.days_to_new_tumor_event_after_initial_treatment,
    survivalNotNas$follow_ups.follow_up.3.days_to_new_tumor_event_after_initial_treatment,
    survivalNotNas$follow_ups.follow_up.4.days_to_new_tumor_event_after_initial_treatment,
    survivalNotNas$follow_ups.follow_up.days_to_new_tumor_event_after_initial_treatment,
    survivalNotNas$new_tumor_events.new_tumor_event.days_to_new_tumor_event_after_initial_treatment),
    1,max))
#############################################
df<-data.frame(survivalNotNas$daysRFS,SUMNASRFS) 
survivalNotNas$status[survivalNotNas$days_to_death==0]<-0
#############################################
# Be careful here is the definition
# days are defined as RFS days
survivalNotNas$days<-survivalNotNas$daysRFS
survivalNotNas$days[survivalNotNas$daysRFS==0]<-
  survivalNotNas$daysOS[survivalNotNas$daysRFS==0]
#########################################################
#optimo
survivalNotNas$group<-rep("g0", nrow(survivalNotNas))
survivalNotNas$group[survivalNotNas$psi>50.83]<-"g1"
#########################################################
#here is OS
fit <- survfit(Surv(daysOS/30, status) ~ group,  
               data = survivalNotNas)
#########################################################
#sum(is.infinite(survivalNotNas$daysOS))#check
##################################
```

### OS and RF Survival plots using optimo PSI (according psichomics: 50.83)
Overall Survival (OS) 
The event call is derived from "vital status" parameter. 
The time_to_event is in days, equals to days_to_death if patient deceased.
In the case of a patient is still living, the time variable is the maximum(days_to_last_known_alive, days_to_last_followup).  This pair of clinical parameters are called _EVENT and _TIME_TO_EVENT on the cancer browser. 

In our case, we use these columns:
  + days_to_death, 
  + days_to_last_followup,
  + days_to_last_known_alive

* Recurrence Free Survival (RFS)
The event call is derived from "new_tumor_event_after_initial_treatment" parameter. The time_to_event is in days, equals to max (days_to_new_tumor_event_after_initial_treatment, days_to_tumor_recurrence)  if there is an event. In the case of no event, the time variable is time of overall survival. The pair of clinical parameters are called _RFS and _RFS_IND on the cancer browser. 

Source: <https://groups.google.com/forum/#!topic/ucsc-cancer-genomics-browser/YvKnWZSsw1Q>
  
  
```{r makeNotNAs2, message=FALSE, warning=FALSE}
title<-paste("Overall Surv. PSI 50.83")
ggsurvplot(fit, 
           data = survivalNotNas,
           conf.int = F,
           pval = TRUE,
           fun = "pct",
           risk.table = TRUE,
           size = 1,
           palette = c("skyblue","tomato"),
           legend = "right",
           legend.labs = c("g0","g1"),
           title=title)
```

```{r firNotNas, echo = FALSE, message=FALSE, warning=FALSE}
fit <- survfit(Surv(days/30, status) ~ group, 
               data = survivalNotNas)
title<-paste("Rec Free Surv PSI 50.83")

ggsurvplot(fit, 
           data = survivalNotNas,
           conf.int = F,
           pval = TRUE,
           fun = "pct",
           risk.table = TRUE,
           size = 1,
           palette = c("skyblue","tomato"),
           legend = "right",
           legend.labs = c("g0","g1"),
           title=title)
```

### Compute optimo PSI inside G0 (psi  <  50.83) 

```{r optG0, echo = FALSE, message=FALSE, warning=FALSE}
clinicalTumor<-data.frame(days_to_last_follow=finalV$days_to_last_follow, 
                          days_to_death=finalV$days_to_death,
                          stage=finalV$stage,
                          gender=finalV$gender,
                          psi=finalV$psi)
clinicalTumor<-clinicalTumor[clinicalTumor$psi<50.83,]
names(clinicalTumor) <- c("patient.days_to_last_followup", 
                          "patient.days_to_death",
                          "patient.stage_event.pathologic_stage",
                          "patient.gender",
                          "psi")
timeStart  <- "days_to_death"
event      <- "days_to_death"
psi<-as.numeric(clinicalTumor$psi)
opt <- optimalSurvivalCutoff(clinicalTumor,
                             psi, censoring="right", 
                             event="days_to_death", 
                             timeStart="days_to_death")
(optimalCutoff <- opt$par)    # Optimal exon inclusion level 50.83
label     <- labelBasedOnCutoff(psi, round(optimalCutoff, 2), 
                                label="PSI values")
survTerms <- processSurvTerms(clinicalTumor, censoring="right",
                              event="days_to_death", timeStart="days_to_death",
                              group=label, scale="months")
surv <- survfit(survTerms)
pvalue <- testSurvival(survTerms)
plotSurvivalCurves(surv, pvalue=pvalue, mark=FALSE)
```

#### OS and RFS in  G0 using optimo psi 15.93

```{r optG0OS, echo = FALSE, message=FALSE, warning=FALSE}
survivalLow<-survivalNotNas[survivalNotNas$group=="g0",]
survivalLow$group<-rep("g0", nrow(survivalLow))
survivalLow$group[survivalLow$psi>15.93]<-"g1"
fit <- survfit(Surv(daysOS/30, status) ~ group,  
               data = survivalLow)
title<-paste("OS PSI 15.93  (optimal psi inside g0 Low)")
ggsurvplot(fit, 
           data = survivalLow,
           conf.int = F,
           pval = TRUE,
           fun = "pct",
           risk.table = TRUE,
           size = 1,
           palette = c("skyblue","tomato"),
           legend = "right",
           legend.labs = c("g0","g1"),
           title=title)
```

```{r optG0RFS, echo = FALSE,message=FALSE, warning=FALSE}
fit <- survfit(Surv(days/30, status) ~ group,  
               data = survivalLow)
title<-paste("RFS PSI 15.93  (optimal psi inside g0 Low)")
ggsurvplot(fit, 
           data = survivalLow,
           conf.int = F,
           pval = TRUE,
           fun = "pct",
           risk.table = TRUE,
           size = 1,
           palette = c("skyblue","tomato"),
           legend = "right",
           legend.labs = c("g0","g1"),
           title=title)
```

### Compute optimo PSI inside G1 (psi >50.83) 

```{r optG1, echo = FALSE, message=FALSE, warning=FALSE}
clinicalTumor<-data.frame(days_to_last_follow=finalV$days_to_last_follow, 
                          days_to_death=finalV$days_to_death,
                          stage=finalV$stage,
                          gender=finalV$gender,
                          psi=finalV$psi)
clinicalTumor<-clinicalTumor[clinicalTumor$psi>=50.83,]
names(clinicalTumor) <- c("patient.days_to_last_followup", 
                          "patient.days_to_death",
                          "patient.stage_event.pathologic_stage",
                          "patient.gender",
                          "psi")
timeStart  <- "days_to_death"
event      <- "days_to_death"
psi<-as.numeric(clinicalTumor$psi)
opt <- optimalSurvivalCutoff(clinicalTumor,
                             psi, censoring="right", 
                             event="days_to_death", 
                             timeStart="days_to_death")
(optimalCutoff <- opt$par)    # 
(optimalPvalue <- opt$value)  # 
label     <- labelBasedOnCutoff(psi, round(optimalCutoff, 2), 
                                label="PSI values")
survTerms <- processSurvTerms(clinicalTumor, censoring="right",
                              event="days_to_death", timeStart="days_to_death",
                              group=label, scale="months")
surv <- survfit(survTerms)
pvalue <- testSurvival(survTerms)
plotSurvivalCurves(surv, pvalue=pvalue, mark=FALSE)
```

#### OS and RFS in  G1 using optimo psi 64.11

```{r optG1OS, echo = FALSE, message=FALSE, warning=FALSE}
survivalHigh<-survivalNotNas[survivalNotNas$group=="g1",]
survivalHigh$group<-rep("g0", nrow(survivalHigh))
survivalHigh$group[survivalHigh$psi>64.11]<-"g1"
fit <- survfit(Surv(daysOS/30, status) ~ group,  
               data = survivalHigh)
title<-paste("OS PSI 64.11  (optimal psi inside g1 high)")
ggsurvplot(fit, 
           data = survivalHigh,
           conf.int = F,
           pval = TRUE,
           fun = "pct",
           risk.table = TRUE,
           size = 1,
           palette = c("skyblue","tomato"),
           legend = "right",
           legend.labs = c("g0","g1"),
           title=title)
```

```{r optG1RFS, echo = FALSE,message=FALSE, warning=FALSE}
fit <- survfit(Surv(days/30, status) ~ group,  
               data = survivalHigh)
title<-paste("RFS PSI 64.11  (optimal psi inside g1 high)")
ggsurvplot(fit, 
           data = survivalHigh,
           conf.int = F,
           pval = TRUE,
           fun = "pct",
           risk.table = TRUE,
           size = 1,
           palette = c("skyblue","tomato"),
           legend = "right",
           legend.labs = c("g0","g1"),
           title=title)
```

## (2nd strategy) Scan pvalue across PSI values (whole dataset)

```{r FindOptimalPSI, echo = FALSE,message=FALSE, warning=FALSE}
survPSI<-survivalNotNas
df<-data.frame(matrix(NA, 
                      ncol=2, 
                      nrow= length(seq(min(survPSI$psi),max(survPSI$psi), by=0.05))))
colnames(df)<-c("psi","pval")
values<-seq(min(survPSI$psi),
            max(survPSI$psi), by=0.05)

for (i in 1:length(values) ){
  df[i,1]<-values[i]
  psi=values[i]
  survPSI$group<-rep("g0", nrow(survPSI))
  survPSI$group[survPSI$psi>psi]<-"g1"
  fit <- survfit(Surv(daysOS/30, status) ~ group,  data = survPSI)
  pval<-surv_pvalue(fit, data = NULL, method = "survdiff",
                    test.for.trend = FALSE, combine = FALSE)[2]
  df[i,2]<-pval
}
title<-paste(cancertype, "PSI OPTIMO",df[which.min(df$pval),1], "pval", round(df[which.min(df$pval),2], digits = 4))
ggplot(data=df, 
       aes(x=psi, y=pval, group=1)) + 
  geom_line(colour="grey")+
  theme_bw()+
  ggtitle(title)+
  geom_hline(yintercept=range(0.05, 0.1), color='coral',       size=0.5)
```

### OS and RFS plots **whole dataset** using optimo PSI (18.28)

```{r FindOptimalPSIb, echo = FALSE,message=FALSE, warning=FALSE}
survPSI$group<-rep("g0", nrow(survPSI))
survPSI$group[survPSI$psi>=18.28]<-"g1"
################################################################
fit <- survfit(Surv(daysOS/30, status) ~ group,  data = survPSI)
title<-paste("Optimo 18.28 PSI OS")
ggsurvplot(fit, 
           data = survPSI,
           conf.int = F,
           pval = TRUE,
           fun = "pct",
           risk.table = TRUE,
           size = 1,
           palette = c("skyblue","tomato"),
           legend = "right",
           legend.labs = c("g0","g1"),
           title=title)
```

```{r FindOptimalPSIbRFS, echo = FALSE,message=FALSE, warning=FALSE}
fit <- survfit(Surv(days/30, status) ~ group,  data = survPSI)
title<-paste("Optimo 18.28 PSI RFS")
ggsurvplot(fit, 
           data = survPSI,
           conf.int = F,
           pval = TRUE,
           fun = "pct",
           risk.table = TRUE,
           size = 1,
           palette = c("skyblue","tomato"),
           legend = "right",
           legend.labs = c("g0","g1"),
           title=title)
```

### Scan for optimal in G0 (psi <18.28)

```{r FindOptimalPSIe, echo = FALSE,message=FALSE, warning=FALSE}
survPSI<-survivalNotNas[survivalNotNas$psi<=18.28,]
df<-data.frame(matrix(NA, 
                      ncol=2, 
                      nrow= length(seq(min(survPSI$psi),max(survPSI$psi), by=0.05))))
colnames(df)<-c("psi","pval")
values<-seq(min(survPSI$psi),max(survPSI$psi), by=0.05)

for (i in 1:length(values) ){
  df[i,1]<-values[i]
  psi=values[i]
  survPSI$group<-rep("g0", nrow(survPSI))
  survPSI$group[survPSI$psi>psi]<-"g1"
  fit <- survfit(Surv(daysOS/30, status) ~ group,  data = survPSI)
  pval<-surv_pvalue(fit, data = NULL, method = "survdiff",
                    test.for.trend = FALSE, combine = FALSE)[2]
  df[i,2]<-pval
}

title<-paste(cancertype, "PSI OPTIMO G0",df[which.min(df$pval),1], "pval", round(df[which.min(df$pval),2], digits = 4))

ggplot(data=df, 
       aes(x=psi, y=pval, group=1)) + 
  geom_line(colour="grey")+
  theme_bw()+
  ggtitle(title)+
  geom_hline(yintercept=range(0.05, 0.1), color='coral',       size=0.5)
```

```{r FindOptimalPSIf, echo = FALSE,message=FALSE, warning=FALSE}
survPSI$group<-rep("g0", nrow(survPSI))
survPSI$group[survPSI$psi>=11.88]<-"g1"
################################################################
fit <- survfit(Surv(daysOS/30, status) ~ group,  data = survPSI)
title<-paste("Optimo G0 11.88 PSI OS")
ggsurvplot(fit, 
           data = survPSI,
           conf.int = F,
           pval = TRUE,
           fun = "pct",
           risk.table = TRUE,
           size = 1,
           palette = c("skyblue","tomato"),
           legend = "right",
           legend.labs = c("g0","g1"),
           title=title)
```

```{r FindOptimalPSIfRFS, echo = FALSE,message=FALSE, warning=FALSE}
################################################################
fit <- survfit(Surv(days/30, status) ~ group,  data = survPSI)
title<-paste("Optimo G0 11.88 PSI RFS")
ggsurvplot(fit, 
           data = survPSI,
           conf.int = F,
           pval = TRUE,
           fun = "pct",
           risk.table = TRUE,
           size = 1,
           palette = c("skyblue","tomato"),
           legend = "right",
           legend.labs = c("g0","g1"),
           title=title)
```

### Scan for optimal in G1 (psi >18.28)

```{r FindOptimalPSIeG1, echo = FALSE,message=FALSE, warning=FALSE}
survPSI<-survivalNotNas[survivalNotNas$psi>=18.28,]
df<-data.frame(matrix(NA, 
                      ncol=2, 
                      nrow= length(seq(min(survPSI$psi),max(survPSI$psi), by=0.05))))
colnames(df)<-c("psi","pval")
values<-seq(min(survPSI$psi),max(survPSI$psi), by=0.05)

for (i in 1:length(values) ){
  df[i,1]<-values[i]
  psi=values[i]
  survPSI$group<-rep("g0", nrow(survPSI))
  survPSI$group[survPSI$psi>psi]<-"g1"
  fit <- survfit(Surv(daysOS/30, status) ~ group,  data = survPSI)
  pval<-surv_pvalue(fit, data = NULL, method = "survdiff",
                    test.for.trend = FALSE, combine = FALSE)[2]
  df[i,2]<-pval
}

title<-paste(cancertype, "PSI OPTIMO G1",df[which.min(df$pval),1], "pval", round(df[which.min(df$pval),2], digits = 4))

ggplot(data=df, 
       aes(x=psi, y=pval, group=1)) + 
  geom_line(colour="grey")+
  theme_bw()+
  ggtitle(title)+
  geom_hline(yintercept=range(0.05, 0.1), color='coral',       size=0.5)
```

```{r FindOptimalPSIfG1, echo = FALSE,message=FALSE, warning=FALSE}
survPSI$group<-rep("g0", nrow(survPSI))
survPSI$group[survPSI$psi>=21.15]<-"g1"
################################################################
fit <- survfit(Surv(daysOS/30, status) ~ group,  data = survPSI)
title<-paste("Optimo G1 21.15 PSI OS")
ggsurvplot(fit, 
           data = survPSI,
           conf.int = F,
           pval = TRUE,
           fun = "pct",
           risk.table = TRUE,
           size = 1,
           palette = c("skyblue","tomato"),
           legend = "right",
           legend.labs = c("g0","g1"),
           title=title)
```
```{r FindOptimalPSIRFSG1, echo = FALSE,message=FALSE, warning=FALSE}
################################################################
fit <- survfit(Surv(days/30, status) ~ group,  data = survPSI)
title<-paste("Optimo G1 21.15 PSI RFS")
ggsurvplot(fit, 
           data = survPSI,
           conf.int = F,
           pval = TRUE,
           fun = "pct",
           risk.table = TRUE,
           size = 1,
           palette = c("skyblue","tomato"),
           legend = "right",
           legend.labs = c("g0","g1"),
           title=title)
```

## New analysis 25.05

Cuando separamos los LUAD TCGA patients por supervivencia hacemos dos grupos >0.5, donde los pacientes mueren de forma más temprana. Sin embargo, en este grupo me gustaría ver algunos más detalles, podríamos ver la distribución, dentro de ese grupo por: 
* genero, 
* estadio y 
* mutaciones en los tumores.
A ver si los pacientes que tienen PSI >0.5 tienen más cosas en común.

```{r moreG1, echo = FALSE,message=FALSE, warning=FALSE}
table(survivalNotNas$group)
survivalHigh<-survivalNotNas[survivalNotNas$group=="g1",]
dim(survivalHigh)
```

### Divide G1 By gender, make g0=MALE, g1=FEMALE:

```{r moreG1gender, echo = FALSE,message=FALSE, warning=FALSE}
table(survivalHigh$gender)
survivalHigh$group[survivalHigh$gender=="male"]<-"g0"
survivalHigh$group[survivalHigh$gender=="female"]<-"g1"
table(survivalHigh$group)
```

```{r g1GenderOS, echo = FALSE,message=FALSE, warning=FALSE}
################################################################
fit <- survfit(Surv(daysOS/30, status) ~ group,  data = survivalHigh)
title<-paste("G1 Gender PSI OS")
ggsurvplot(fit, 
           data = survivalHigh,
           conf.int = F,
           pval = TRUE,
           fun = "pct",
           risk.table = TRUE,
           size = 1,
           palette = c("skyblue","tomato"),
           legend = "right",
           legend.labs = c("g0","g1"),
           title=title)
```

```{r g1GenderRFS, echo = FALSE,message=FALSE, warning=FALSE}
################################################################
fit <- survfit(Surv(days/30, status) ~ group,  data = survivalHigh)
title<-paste("G1 Gender PSI RFS")
ggsurvplot(fit, 
           data = survivalHigh,
           conf.int = F,
           pval = TRUE,
           fun = "pct",
           risk.table = TRUE,
           size = 1,
           palette = c("skyblue","tomato"),
           legend = "right",
           legend.labs = c("g0","g1"),
           title=title)
```

### Divide G1 By AggStage, G0=early, G1=late:
```{r moreG1AggStage, echo = FALSE,message=FALSE, warning=FALSE}
table(survivalHigh$aggStage)
survivalHigh$group[survivalHigh$aggStage=="early"]<-"g0"
survivalHigh$group[survivalHigh$aggStage=="late"]<-"g1"
table(survivalHigh$group)
```

```{r g1AggStageOS, echo = FALSE,message=FALSE, warning=FALSE}
################################################################
fit <- survfit(Surv(daysOS/30, status) ~ group,  data = survivalHigh)
title<-paste("G1 AggStage PSI OS")
ggsurvplot(fit, 
           data = survivalHigh,
           conf.int = F,
           pval = TRUE,
           fun = "pct",
           risk.table = TRUE,
           size = 1,
           palette = c("skyblue","tomato"),
           legend = "right",
           legend.labs = c("g0","g1"),
           title=title)
```

```{r g1GAggStageRFS, echo = FALSE,message=FALSE, warning=FALSE}
################################################################
fit <- survfit(Surv(days/30, status) ~ group,  data = survivalHigh)
title<-paste("G1 AggStage PSI RFS")
ggsurvplot(fit, 
           data = survivalHigh,
           conf.int = F,
           pval = TRUE,
           fun = "pct",
           risk.table = TRUE,
           size = 1,
           palette = c("skyblue","tomato"),
           legend = "right",
           legend.labs = c("g0","g1"),
           title=title)
```

### Check Mutations: in this small group (G1)

```{r moreG1egfr, echo = FALSE,message=FALSE, warning=FALSE}
print("EGFR performed") 
print(table(survivalHigh$egfr_mutation_performed))
#kruskal.test(finalV$psi ~ finalV$egfr_mutation_performed)
print("EGFR results") 
print(table(survivalHigh$egfr_mutation_result))

print("kras mutation found") 
print(table(survivalHigh$kras_mutation_found))

print("kras mutation result") 
print(table(survivalHigh$kras_mutation_result))
```

## Gender analysis whole dataset
Viendo las diferencias significativas de NUMB exon 9 PSI male vs female. Me preguntaba si crees que esto podría tener un efecto en las gráficas de supervivencia.


```{r GenderWhole, echo = FALSE,message=FALSE, warning=FALSE}
survivalNotNas$group<-rep("g0", nrow(survivalNotNas))
survivalNotNas$group[survivalNotNas$gender=="female"]<-"g1"
table(survivalNotNas$gender)
survivalNotNas$group[survivalNotNas$gender=="male"]<-"g0"
survivalNotNas$group[survivalNotNas$gender=="female"]<-"g1"
table(survivalNotNas$group)

fit <- survfit(Surv(daysOS/30, status) ~ group,  data = survivalNotNas)
title<-paste("Whole dataset Gender PSI OS")
ggsurvplot(fit, 
           data = survivalNotNas,
           conf.int = F,
           pval = TRUE,
           fun = "pct",
           risk.table = TRUE,
           size = 1,
           palette = c("skyblue","tomato"),
           legend = "right",
           legend.labs = c("g0","g1"),
           title=title)
```

```{r WholeGenderRFS, echo = FALSE,message=FALSE, warning=FALSE}
################################################################
fit <- survfit(Surv(days/30, status) ~ group,  data = survivalNotNas)
title<-paste("Whole Gender PSI RFS")
ggsurvplot(fit, 
           data = survivalNotNas,
           conf.int = F,
           pval = TRUE,
           fun = "pct",
           risk.table = TRUE,
           size = 1,
           palette = c("skyblue","tomato"),
           legend = "right",
           legend.labs = c("g0","g1"),
           title=title)
```


